Using spatially biased variables in ecosystem condition accounting with a GIS based workflow

### 2024-01-19 manuscript/HIAs.qmd
kart

@@

| Variable code        | Variable name                                    | Measurement unit                                                                        | Description                                                                                                    | Reference                          |
|--------------|--------------|--------------|----------------|--------------|
| 7FK                  | Prevalence of alien species                      | Unitless, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                             | @runehalvorsen2019                 |
| 7SE                  | Human caused abration or abration-caused erosion | Unitless, ordinal, non-linear scale from 1 to 4.                                        | Measures the frequency of imagined 4 m^2^ quadrats layed over the area that has some sign of abration          | @runehalvorsen2019                 |
| PRSL                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.                                        | Same as 7SE, but recorded at a higher resolution                                                               | @miljødirektoratet                 |
| 7TK                  | Tracks from large vehicles                       | Unitless, ordinal, non-linear scale from 1 to 4.                                        | Measures the frequency of imagined 100 m^2^ quadrats layed over the area that has some signs of vehicle tracks | @rune_halvorsen_dokumentasjon_2019 |
| PRTK                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.                                        | Same as 7TK, but recorded at a higher resolution                                                               | @miljødirektoratet                 |
| Infrastructure Index | Infrastructure Index                             | Unitelss linear scale from 0 to 13.2                                                    | Unitelss index ranging from from 0 to 13.2                                                                     | @erikstad_index_2023               |
| HIA                  | Human Impact area                                | Ordinal, non-linear scale from 1 to 4                                                   | A categorical representation of the infrastructure index                                                       | this paper                         |

@@

![Schematic workflow followed in this paper. Sub-index A is calculated using spatially explicit aggregation (i.e .a simple overlay taking the mean value).](../images/workflow.jpg){#fig-workflow}

@@

```{r prepData}
# Set up conditional file paths
# The data is too big to be stored on GitHub
dir <- substr(getwd(), 1,2)

naturetypes <- naturetypes |>
  # keep only wetlands and our selected nature types
  filter(hovedøkosystem == "våtmark",
         naturtype %in% c(advsNTs, alienNTS)) |>
  # calculate the areas (m2) of the polygons
  mutate(area = SHAPE |> st_area()) |>
  # the variable codes and values are all in the same column
  separate_rows(ninBeskrivelsesvariable, sep=",") |>
  separate(col = ninBeskrivelsesvariable,
           into = c("NiN_variable_code", "NiN_variable_value"),
           sep = "_",
           remove=F) |>
  mutate(NiN_variable_value = as.numeric(NiN_variable_value)) |>
  filter(NiN_variable_code %in% c(myVars, myVars2)) |>
  select(id = identifikasjon_lokalId,
         municipality = kommunenummer,
         year = kartleggingsår,
         mosaic = mosaikk,
         quality = lokalitetskvalitet,
         biodiversity = naturmangfold,
         condition = tilstand,
         natureType = naturtype,
         variable = NiN_variable_code,
         value = NiN_variable_value,
         area)

```{r getNorwayOutline}
outline <- sf::read_sf("../data/outlineOfNorway_EPSG25833.shp")
```

# find the correct layer
#st_layers(path)
muni <- sf::read_sf(path, options = "ENCODING=UTF8", layer = "kommune")
#st_crs(muni) #25833

# This returns some multisurfaces as well
table(st_geometry_type(muni))

muni <- ensure_multipolygons(muni)
#> Writing layer `file34a3261c663332' to data source 
#>   `/tmp/RtmpkT1dSD/file34a3261c663332.gpkg' using driver `GPKG'
#> Writing 363 features with 13 fields and geometry type Unknown (any).
#> Reading layer `file34a3261c663332' from data source 
#>   `/tmp/RtmpkT1dSD/file34a32673f6fef.gpkg' 
#>   using driver `GPKG'
#> Simple feature collection with 363 features and 13 fields
#> Geometry type: MULTIPOLYGON
#> Dimension:     XY
#> Bounding box:  xmin: -99551.21 ymin: 6426048 xmax: 1121941 ymax: 7962744
#> Projected CRS: ETRS89 / UTM zone 33N
table(st_geometry_type(muni)) #OK

# code to show all municipalities
#tm_shape(muni)+
#  tm_polygons(col="kommunenummer")+
#  tm_layout(legend.outside = T)

# Transform to match the EDM (see below)
muni <- muni |>
  st_transform(25832)

# subset of the three taregt municipalities
muni3 <- muni |>
  filter(kommunenummer %in% c("3020", #Nordre Follo
                              "5001", # Trondheim
                              "5418"  # Målselv
                              ))
# To crop EDM, I need the three municipalities seprately. 
nf <- muni3 |>
  filter(kommunenummer == "3020")
tr <- muni3 |>
  filter(kommunenummer == "5001")
me <- muni3 |>
  filter(kommunenummer == "5418")

# I need to intersect nature types with municipality
# and also to get the dat coverage polygon
tm_shape(muni3)+
  tm_polygons()+
  tm_facets(by = "kommunenummer")+
  tm_shape(naturetypes)+
  tm_polygons()
```

>>>>

```{r prepData, eval=F}
# Set up conditional file paths
# The data is too big to be stored on GitHub
dir <- substr(getwd(), 1,2)

naturetypes <- naturetypes |>
  # keep only wetlands and our selected nature types
  filter(hovedøkosystem == "våtmark",
         naturtype %in% c(advsNTs, alienNTS)) |>
  # calculate the areas (m2) of the polygons
  mutate(area = SHAPE |> st_area()) |>
  # the variable codes and values are all in the same column
  separate_rows(ninBeskrivelsesvariable, sep=",") |>
  separate(col = ninBeskrivelsesvariable,
           into = c("NiN_variable_code", "NiN_variable_value"),
           sep = "_",
           remove=F) |>
  mutate(NiN_variable_value = as.numeric(NiN_variable_value)) |>
  filter(NiN_variable_code %in% c(myVars, myVars2)) |>
  select(id = identifikasjon_lokalId,
         municipality = kommunenummer,
         year = kartleggingsår,
         mosaic = mosaikk,
         quality = lokalitetskvalitet,
         biodiversity = naturmangfold,
         condition = tilstand,
         natureType = naturtype,
         variable = NiN_variable_code,
         value = NiN_variable_value,
         area) |>
  st_transform(25832)

```{r}
naturetypes <- readRDS("../data/surveyData.RDS")
```

```{r getNorwayOutline}
outline <- sf::read_sf("../data/outlineOfNorway_EPSG25833.shp") |>
  st_transform(25832)
```

# find the correct layer
#st_layers(path)
muni <- sf::read_sf(path, layer = "kommune") |>
  st_transform(25832)

# This returns some multisurfaces as well
#table(st_geometry_type(muni))

muni <- ensure_multipolygons(muni)
#table(st_geometry_type(muni)) #OK

# subset of the three target municipalities
muni3 <- muni |>
  filter(kommunenummer %in% c("3020", #Nordre Follo
                              "5001", # Trondheim
                              "5418"  # Målselv
                              )) |>
  mutate(Municipality = case_when(
    kommunenummer == "3020" ~ "Nordre Follo",
    kommunenummer == "5001" ~ "Trondheim",
    kommunenummer == "5418" ~ "Målselv"
  ))
# To crop EDM, I need the three municipalities seprately. 
nf <- muni3 |>
  filter(kommunenummer == "3020")
tr <- muni3 |>
  filter(kommunenummer == "5001")
me <- muni3 |>
  filter(kommunenummer == "5418")

# I need to intersect nature types with municipality
nature3 <- naturetypes |>
  st_intersection(muni3)

# and also to get the data coverage polygon
dk <-   read_sf("../data/dekningskartNF/Naturtyper NiN - Dekningskart.shp") |>
  rbind(read_sf("../data/dekningskartTR/Naturtyper NiN - Dekningskart.shp")) |>
  rbind(read_sf("../data/dekningskartME/Naturtyper NiN - Dekningskart.shp")) |>
  st_transform(st_crs(muni3)) |>
  st_intersection(muni3) |>
  st_union()

outline2 <- outline |>
  st_intersection(muni3)

ocean <- muni3 |>
  st_difference(outline)

tm_shape(muni3)+
    tm_borders()+
  tm_facets(by = "Municipality")+
  tm_shape(outline2)+
    tm_fill(col = "lightgreen",
             alpha=.4)+
  tm_shape(ocean)+
    tm_polygons(col = "lightblue",
                border.col = "black",
            alpha=1)+
  #tm_shape(muni3)+
  #  tm_borders()+
  tm_shape(dk)+
    tm_polygons(col = "grey",
                alpha=.8)+
  tm_shape(nature3)+
  tm_polygons(col = "red",
              border.col = "red",
              lwd=3)

@@

| Variable code        | Variable name                                    | Measurement unit                                                                        | Description                                                                                                    | Reference                          |
|---------------|---------------|---------------|---------------|---------------|
| 7FK                  | Prevalence of alien species                      | Unitless, ordinal, non-linear scale from 1 (no alien species) to 7 (only alien species) | The fraction of the species composition made up from alien species                                             | @runehalvorsen2019                 |
| 7SE                  | Human caused abration or abration-caused erosion | Unitless, ordinal, non-linear scale from 1 to 4.                                        | Measures the frequency of imagined 4 m^2^ quadrats layed over the area that has some sign of abration          | @runehalvorsen2019                 |
| PRSL                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.                                        | Same as 7SE, but recorded at a higher resolution                                                               | @miljødirektoratet                 |
| 7TK                  | Tracks from large vehicles                       | Unitless, ordinal, non-linear scale from 1 to 4.                                        | Measures the frequency of imagined 100 m^2^ quadrats layed over the area that has some signs of vehicle tracks | @rune_halvorsen_dokumentasjon_2019 |
| PRTK                 | *as above*                                       | Unitless, ordinal, non-linear scale from 0 to 7.                                        | Same as 7TK, but recorded at a higher resolution                                                               | @miljødirektoratet                 |
| Infrastructure Index | Infrastructure Index                             | Unitelss linear scale from 0 to 13.2                                                    | Unitelss index ranging from from 0 to 13.2                                                                     | @erikstad_index_2023               |
| HIA                  | Human Impact area                                | Ordinal, non-linear scale from 1 to 4                                                   | A categorical representation of the infrastructure index                                                       | this paper                         |

@@

![Schematic workflow followed in this paper. Sub-index A is calculated using spatially explicit aggregation (i.e .a simple overlay taking the mean value).](../images/workflow.jpg){#fig-workflow alt="Schematic workflow followed in this paper. Sub-index A is calculated using spatially explicit aggregation (i.e .a simple overlay taking the mean value)."}

@@
